---
topic: "land"
title: "Roads & Roadless Areas in British Columbia"
output: envreportutils.internal::print_ver
---

<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(png)
library(here)
library(readr)
library(dplyr)
library(rmapshaper)
library(bcmaps)
library(ggplot2)
library(sf)
library(patchwork)
library(purrr)
library(forcats)


opts_chunk$set(echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, error=FALSE)

proj_path <- here("out/")
rds_path <- here("tmp/")

## Load csv for road summary and list of plots
plot_list <- readRDS(paste0(rds_path, "plotlist.rds"))


## axis text theme
theme_text <- theme(axis.text = element_text(size=10),
               axis.title = element_text(size=10))

```

While roads serve the interests of the public, First Nations, and
industries, they are also one of the most widespread forms of landscape
change^[1](#fn1)^. Roads can threaten biological diversity and ecosystem
health through habitat loss and fragmentation, expanding development,
spread of [invasive
species](http://www.env.gov.bc.ca/soe/indicators/plants-and-animals/invasive-species.html)
and increasing human-wildlife interactions—for example animal mortality
from collisions with vehicles^[1](#fn1)^. This indicator reports on the
amount and distribution of roads and roadless areas in British Columbia.

-   **British Columbia has approximately 719,000 kilometres of roads.**
    Roads include all paved roads and all unpaved roads and trails that allow motorized vehicle use.
-   **Unpaved roads make up 92 percent of the roads in British
    Columbia.** B.C. has a long history of natural resource use
    resulting in this large gravel road network^[2](#fn2)^.
-   **Two thirds of British Columbia's landbase is roadless.** The
    amount and distribution of roadless area&mdash;areas further than 500
    metres from a road&mdash;varies considerably across British Columbia.

Large areas that are not influenced by roads or other landscape
disturbances provide benefits and opportunities for British Columbia and
British Columbians. These large, roadless&mdash;or intact&mdash;areas provide
conservation benefits, for example providing habitat for species such as
[grizzly
bear](http://www.env.gov.bc.ca/soe/indicators/plants-and-animals/grizzly-bears.html),
maintaining ecosystem services, such as water availability and flood
control, and supporting adaptation to climate change. They also provide
economic, recreational and wilderness opportunities.


\begin{center}
\fbox{
  \begin{minipage}[c]{14cm}
  
   \begingroup\centering{\textbf{B.C. Digital Road Atlas}}
   
     \endgroup
     
    \vspace{1mm}
    
    \text{The Digital Road Atlas is the best available single source of road data for the Province of B.C.}
    
       \vspace{2mm}
    
    \text{The B.C. Government is working toward updating and integrating all resource roads into this}
    \text{this single comprehensive database.}
    
       \vspace{2mm}
    
    \href{https://www2.gov.bc.ca/gov/content?id=21FFEC94B0AD40818D2D2AF06D522714}{Explore the B.C. Digital Road Atlas to learn more about roads in B.C.}
    
  \end{minipage}
}
\end{center}

\newpage


## Roads in B.C.

-   British Columbia has approximately 719,000 kilometres of roads, when including all
    road types that allow motorized vehicle use.
-   Unpaved roads&mdash;which include all seasonal, loose and rough surfaced roads&mdash;contribute the most (92%) to the total length of roads in B.C.

```{r road-bar-plot, results='asis', fig.height=4, fig.width=6, fig.align='center'}

soe_roads_sum_chart <- readRDS(paste0(rds_path, "soe_roads_sum_chart.rds"))

barchart <- soe_roads_sum_chart +
  theme_text
plot(barchart)
```

```{r road-map, fig.align='center', out.width="70%"}

road_map <- "soe_roads_map.png"
include_graphics(paste0(proj_path, road_map))
```

\newpage

## Roadless Areas in B.C.

-   Sixty-six percent&mdash;or 62,600,000 hectares&mdash;of British Columbia's
    landbase is roadless, defined as more than 500 metres from a road.
    
```{r roadless-area-figures, results='asis', fig.height=4, fig.width=6, fig.align='center'}

## Filter list of plots for the province plots only
prov_list <- plot_list[names(plot_list) == "Province"]

iwalk(prov_list,  ~ {
  pbc <- .x$barchart + theme_text
  plot(pbc)
  plot(.x$map)
})

```

\newpage

## Roadless Areas within B.C.'s Ecoregions

-   The amount and distribution of roadless areas vary considerably
    across B.C. Twenty seven of the 43 terrestrial
    [ecoregions](https://www2.gov.bc.ca/gov/content?id=80F98E32A7FC45B2AC8F838041CB98A9)
    in B.C. are more than 50 percent roadless, while 6 ecoregions have
    less that 25 percent roadless area.
-   As expected, ecoregions with high [population
    densities](http://www.env.gov.bc.ca/soe/indicators/sustainability/bc-population.html),
    and hence larger road networks, have the lowest amounts of remaining
    roadless area. For example, the Lower Mainland, Georgia-Puget Basin
    and Okanagan Highland ecoregions each have 13 percent roadless
    area (see the Appendix for individual ecoregion results).

\vspace{30pt}

```{r map, warning=FALSE, fig.height=8, fig.align="center"}
ecoreg_summary <- read_csv(paste0(proj_path, "data/bc_roadless_areas_ecoreg_summary.csv")) %>% 
  filter(name != "Province", roaded_class == "Not Roaded") %>% 
  mutate(percent_unroaded = percent_in_distance_class)

ecoreg <- ecoregions() %>% 
  filter(!ECOREGION_CODE %in% c("HCS", "IPS", "OPS", "SBC", "TPC")) %>% 
  ms_simplify() %>% 
  st_intersection(bc_bound()) %>% 
  group_by(ECOREGION_NAME) %>% 
  summarize() %>% 
  left_join(ecoreg_summary, by = c("ECOREGION_NAME" = "name")) %>% 
  ggplot() +
  geom_sf(aes(fill=percent_unroaded, geometry = SHAPE), size = 0.1) +
  geom_sf(data = bc_bound(), fill = NA, size = 0.2) +
  coord_sf(datum = NA) +
#  scale_fill_continuous(name = "% Roadless Area") +
  scale_fill_distiller(name = "% Roadless Area", palette = "YlGn", trans = "reverse",
                       breaks = c(20, 30, 40, 50, 60, 70, 80, 90)) +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.1))
plot(ecoreg)

```

\newpage

## Methods

This indicator is based on the B.C. [Digital Road Atlas](https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e)the best available single source of road data for British Columbia (accessed on May 1st, 2018). All paved roads and unpaved roads and trails that *allow* motorized vehicle use&mdash;as detailed in the [Digital Road Atlas Data Dictionary](https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e)&mdash;were included in the analysis. Some road features available in the [Digital Road Atlas](https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e) were excluded, including non-roads (ferry routes, proposed roads, pedestrian only roads, non-motorized trails and water access routes) and non-active roads (decommissioned and overgrown roads). The indicator presents a coarse-scale snapshot of the amount and distribution of roads in British Columbia&mdash;data completeness, data quality, road condition or intensity of road use were not evaluated. Roadless areas were determined by applying a 500 metre buffer on either side of all roads. A 500 metre buffer is a conservative value for determining influence and has been defined as the minimum zone of influence in a recent assessment of intact landscapes in Canada^[3](#fn3)^. All results are reported to three significant figures. While the indicator results are broadly similar to previous road and roadless area analyses for British Columbia, the results among studies cannot be directly compared due to the use of different road data sources and methodologies^[2,4](#fn4)^.

The [**R**](http://www.r-project.org) code for repeating the analysis and data visualizations presented on this page is available on [GitHub](https://github.com/bcgov/roadless-areas-indicator).

## References and Other Useful Links

-   [B.C. Digital Road Atlas](https://www2.gov.bc.ca/gov/content?id=21FFEC94B0AD40818D2D2AF06D522714)
-   [Biodiversity B.C.](http://www.biodiversitybc.org/EN/main/26.html)
-   ^1^Trombulak, S.C. and C.A. Frissell. 2000. Review of ecological
    effects of roads on terrestrial and aquatic communities.
    Conservation Biology. 14: 18-30. [DOI:
    10.1046/j.1523-1739.2000.99084.x](http://onlinelibrary.wiley.com/doi/10.1046/j.1523-1739.2000.99084.x/full)
-   ^2^[B.C. Forest Practices Board. 2015. Access Management and
    Resource Roads: 2015 Update.
    FPB/SR/49.](https://www.bcfpb.ca/reports-publications/reports/access-management-and-resource-roads-2015-update)
-  ^3^Lee, P.G., M. Hanneman, J.D. Gysbers, R. Cheng and W.
    Smith. 2010. Atlas of Canada’s Intact Forest Landscapes. Edmonton,
    Alberta: Global Forest Watch Canada 10th Anniversary Publication
    \#1. 74 pp.
-   ^4^[British Columbia. Ministry of Environment. 2007. Environmental
    Trends in British Columbia: 2007. Ministry of Environment, State of
    Environment
    Reporting.](https://www2.gov.bc.ca/gov/content?id=07DD1DCE7F784EFEBA5A26C32CB1646D)

## Data

\*By accessing these datasets, you agree to the licence associated with each file, as indicated in parentheses below.

-   [Indicator Summary Data: Amount of Roaded & Roadless Area by Ecoregion in B.C.](https://catalogue.data.gov.bc.ca/dataset/42f52a87-e0a3-4dfa-8a0a-7c0aba69167d) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

-   [Digital Road Atlas (DRA) - Master Partially-Attributed Roads](https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e) (Licence: [B.C. Crown Copyright](https://www2.gov.bc.ca/gov/content?id=1AAACC9C65754E4D89A118B875E0FBDA))
-   [Ecoregions - Ecoregion Ecosystem Classification of British
    Columbia](https://catalogue.data.gov.bc.ca/dataset/d00389e0-66da-4895-bd56-39a0dd64aa78) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (May 2018):
[http://www.env.gov.bc.ca/soe/land/roads.html](http://www.env.gov.bc.ca/soe/land/roads.html)

Email correspondence to: envreportbc@gov.bc.ca

\newpage

## APPENDIX: Roadless Area Summaries for Each Ecoregion in B.C.

\vspace{20pt}

```{r appendix, results='asis', fig.height=4, fig.width=6, fig.align='center'}

## Filter list of plots for the ecoregion plots for the appendix
appendix_list <- plot_list[names(plot_list) != "Province"]

## TESTING
#foo <- appendix_list[1:4] #object for testing iwalk()
##

iwalk(appendix_list,  ~ {
  ebc <- .x$barchart + 
    labs(title = paste0(.y, "\n", "\n", sep = "")) +
    theme(plot.title = element_text(size = 12, face="bold")) +
    theme_text
    plot(ebc)
    plot(.x$map)
   cat("\\newpage ")
})


```


